---
title: "Randomization"
format: html
author: "Charles Appolon"
---

# Introduction  
The goals of this exercise are to:  
- Learn how to randomize a split-plot RCBD in R  
- Learn how to create plot layouts with them  

# Setup  
Here is where we load the packages we will use.  
```{r setup}
# Loading packages
library(tidyverse) # for data wrangling and plotting
library(dplyr)
library(readr)
```

# Treatment design  
Keeping with the same motivational example we've been using, our treatment design is:  

-   2-way factorial **in a split-plot**  
-   **Whole-plot** treatment factor: K fertilizer rates: 0, 30, 60 kg K/ha\
-   **Split-plot** treatment factor: N fertilizer rates: 0, 100, 200 kg N/ha\
-   3 x 3 = **9 treatment combinations** that will be assigned to different sizes of experimental units!  

# Split-plot in RCBD  
## WP randomization  
First, let's randomize the whole-plot treatment factor (K rate) to the blocks.  

I decided to call it block (instead of rep) because the concept of rep in a split-plot depends on which split level (whole-plot or split-plot) we are talking about.  

In a split-plot RCBD, the block is still the level of replication for the whole-plot treatment factor. For the split-plot treatment factor, replication is at the block x WP level.  

Let's define our treatments above in an object.  
```{r}
trts <- crossing(nrate_kgha = c(0, 100, 200),
                 krate_kgha = c(0, 30, 60)
                 ) %>%
  mutate(trtnum = 1:nrow(.)) %>%
  mutate(krate_kgha = paste0(krate_kgha)) %>%
  mutate(trtname1 = paste0(nrate_kgha))
trts
```


```{r wp_rcbd_rand}
# Setting seed to get same randomization result every time
set.seed(930)

wp_rcbd_rand <- crossing(krate_kgha = trts$krate_kgha,
                     rep = 1:4
                     ) %>%
  arrange(rep) %>%
  group_by(rep)%>%
  # Randomizing WP treatment factor to blocks (restricted) 
  mutate(krate_kgha = sample(krate_kgha)) %>%
    # Creating WP plot numbers 
 mutate(wplot = (rep*100)+ 1:3) %>%
  ungroup()
wp_rcbd_rand
```

How many rows do we have in the object above? Why?  
We have 12 rows, as we only randomized the k rate here.
## SP randomization  
Now, let's randomize the split-plot treatment factor (N rate) to the WP treatment plots.  
```{r sp_rcbd_rand}
sp_rcbd_rand <- crossing(
  nrate_kgha = trts$nrate_kgha,    # Rename the column as 'nrate_kgha'
  wplot = wp_rcbd_rand$wplot       # Use the whole plot identifiers from wp_rcbd_rand
) %>%  
  arrange(wplot) %>%    
 # Randomizing SP treatment factor to WP plots (restricted)
   group_by(wplot) %>%
  mutate(nrate_kgha = sample(nrate_kgha)) %>%
  # Creating SP plot numbers 
  mutate(splot = paste0(wplot,"-",1:3)) %>%
  ungroup()
sp_rcbd_rand
```

## Joining WP and SP into one object
```{r spp_rcbd_rand, message=FALSE, warning=FALSE}
spp_rcbd_rand <- wp_rcbd_rand %>%
  right_join(sp_rcbd_rand) %>%
 mutate(realname = paste0(krate_kgha, "+", nrate_kgha))

spp_rcbd_rand
```

How many rows in the object above? Why?
We have 9 rows in block 1. as three levels of N are randomized within 3 levels of K. 
Let's inspect block 1.  
```{r}
spp_rcbd_rand %>%
  filter(rep == 1)
```

Notice wplot and splot columns. What is going on here?  
The wplot does not change for each level of potassium rate. Conversely, splot changes within that wplot for all three levels of nitrogen. This shows that we have 3 whole plots and 9 subplots.

## Plot layout  
```{r spp_rcbd_layout}
spp_rcbd_layout <- spp_rcbd_rand %>%
  # Defining number of rows and columns in the field for plotting purposes
  mutate(row = rep(1:4, each = 9)) %>%
  mutate(col = rep(seq(from = 1, 
                       to = 9, #total number of treatments, 
                       by = 1), 
                   times = 4 #total number of blocks
  )) %>%
  # Creating plot and trtname labels for plotting
  mutate(lab = paste0(splot, "\n", realname ))

spp_rcbd_layout
```

Now, let's plot it.
```{r spp rcbd mapping, message=FALSE, warning=FALSE}
ggplot(spp_rcbd_layout, aes(x = row, 
                                   y = -col)) +
   geom_tile(aes(fill = factor(nrate_kgha)),
            color = "black",
            show.legend = F) +
  geom_tile(aes(color = krate_kgha),
            show.legend = F,
            size = 2,
            linetype = 2,
            fill = NA) +
    scale_fill_manual(values = c("#e5f0db",
                            "#b1d095",
                             "#5e813f")) +
  geom_text(aes(label = splot), fontface = "bold", vjust = -.5, size = 4)+
  geom_text(aes(label = realname), vjust = 1, size = 3)+
  labs(title = "Split-plot RCBD randomization of a\n3 K (whole-plot) x 3 N (split-plot) treatment design") +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5))
```

Notice how:  
  - Each K rate treatment was randomized within each block first, where each WP experimental unit comprises 3 cells. 
  
  - Each N rate treatment was randomized within each WP EU, where each SP experimental unit comprises 1 cell.  

# Exporting to file  
If this is in fact your study, you may want to export both the field randomization table AND the layout map.  

Let's do that below for the RCBD example.  

```{r}
spp_rcbd_rand %>%
  # Rearranging columns  
  dplyr::select(rep, wplot, krate_kgha, 
                splot, nrate_kgha,
                realname) %>%
  write_csv("../output/spp_rcbd_randomization_table.csv")
```

`ggsave()` will save the last plot created in the session, even if from an above chunk. Since the last plot we created was the RCBD layout, we can just go ahead and call ggsave here.  

```{r}
ggsave("../output/spp_rcbd_layoutmap.png",
       width = 6,
       height = 5,
       bg = "white")
```


